---
title: "Surveillance workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Surveillance workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```

## Overview

This vignette demonstrates a complete end-to-end surveillance
analysis: from raw count data to actionable outputs. The workflow
mirrors what a public health genomics team would run weekly.

## Step 1: Load and prepare data

```{r load}
library(lineagefreq)

data(sarscov2_us_2022)
x <- lfq_data(sarscov2_us_2022,
              lineage = variant,
              date    = date,
              count   = count,
              total   = total)
```

## Step 2: Collapse rare lineages

Real surveillance data often contains dozens of low-frequency
lineages. `collapse_lineages()` merges those below a threshold
into an "Other" category.

```{r collapse}
x_clean <- collapse_lineages(x, min_freq = 0.02)
attr(x_clean, "lineages")
```

## Step 3: Fit model

```{r fit}
fit <- fit_model(x_clean, engine = "mlr")
summary(fit)
```

## Step 4: Growth advantages

```{r ga}
ga <- growth_advantage(fit,
                       type = "relative_Rt",
                       generation_time = 5)
ga
```

```{r ga-plot}
autoplot(fit, type = "advantage", generation_time = 5)
```

## Step 5: Identify emerging lineages

```{r emerge}
emerging <- summarize_emerging(x_clean)
emerging[emerging$significant, ]
```

## Step 6: Forecast

```{r forecast}
fc <- forecast(fit, horizon = 28)
autoplot(fc)
```

## Step 7: Assess sequencing needs

How many sequences are needed to reliably detect a lineage
at 2% frequency?

```{r power}
sequencing_power(
  target_precision = 0.05,
  current_freq = c(0.01, 0.02, 0.05, 0.10)
)
```

## Step 8: Extract tidy results

All results are compatible with the broom ecosystem.

```{r tidy}
tidy.lfq_fit(fit)
glance.lfq_fit(fit)
```

## Summary

A typical weekly workflow:

1. `lfq_data()` — ingest new counts
2. `collapse_lineages()` — clean up rare lineages
3. `fit_model()` — estimate dynamics
4. `summarize_emerging()` — flag growing lineages
5. `forecast()` — project forward 4 weeks
6. `autoplot()` — generate report figures
